---
title: "Differential_expression_analysis"
author: "Vikas Kaushik"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---
7. Load relevant R packages.

```{r}
library(ballgown)
library (tidyverse)
library(ggplot2)
library(RSkittleBrewer)
library(genefilter)
library(dplyr)
library(devtools)
library (cowplot)
library(DESeq2)
library(EnhancedVolcano)
library(pheatmap)

```

8. Load the phenotype data for the samples stored in  geuvadis_phenodata.csv

```{r}
# Make sure to chnage the path to 'D:/NGS/chrX_data' where file is located
setwd('D:/NGS/chrX_data')

# Read the phenotype data into an object
geuvadis_phenodata <- read.csv("geuvadis_phenodata.csv")

# View the data frame
class(geuvadis_phenodata)

```

9. Read in the expression data that were calculated by StringTie. Lets's explore a standardized pipeline that can be used to perform standard differential expression analysis.

```{r}
# Make sure geuvadis_phenodata is a valid data frame, Convert it to a data frame if needed
geuvadis_phenodata <- as.data.frame(geuvadis_phenodata)

# Create ballgown object
bg_chrX = ballgown(dataDir = "ballgown", samplePattern = "ERR", pData = geuvadis_phenodata)

```

10. Filter to remove low-abundance genes. 

```{r}
# Filter genes in bg_chrX with row-wise variance greater than 1
bg_chrX_filt <- subset(bg_chrX, "rowVars(texpr(bg_chrX)) > 1", genomesubset = TRUE)

```

11. Identify transcripts that show statistically significant differences between groups. 

```{r}
# Perform differential expression analysis on transcripts, testing for the effect of 'sex' with adjustment for 'population'
results_transcripts = stattest(bg_chrX_filt, feature = "transcript", covariate = "sex", adjustvars = c("population"), getFC = TRUE, meas = "FPKM")

```

12. Identify genes that show statistically significant differences between groups.

```{r}
# Perform differential expression analysis on genes, testing for the effect of 'sex' with adjustment for 'population'
results_genes = stattest(bg_chrX_filt, feature = "gene", covariate = "sex", adjustvars = c("population"), getFC = TRUE, meas = "FPKM")

```

13. Add gene names and gene IDs to the results_transcripts data frame.

```{r}
# Combine results of transcript-level differential expression analysis with gene names and IDs
results_transcripts = data.frame(
  geneNames = ballgown::geneNames(bg_chrX_filt),
  geneIDs = ballgown::geneIDs(bg_chrX_filt),
  results_transcripts
)
```

14. Sort the results from the smallest P value to the largest.

```{r}
# Arrange results_transcripts and results_genes data frames based on p-value
results_transcripts = arrange(results_transcripts, pval)
results_genes = arrange(results_genes, pval)

```

15. Write the results to csv files.

```{r}
# Write transcript-level results to a CSV file
write.csv(results_transcripts, "chrX_transcript_results.csv", row.names = FALSE)

# Write gene-level results to a CSV file
write.csv(results_genes, "chrX_gene_results.csv", row.names = FALSE)

```

16. Identify transcripts and genes with a q value <0.05:

```{r}
# Subset transcript-level results with q-value < 0.05
subset_transcripts = subset(results_transcripts, qval < 0.05)

# Subset gene-level results with q-value < 0.05
subset_genes = subset(results_genes, qval < 0.05)

```

Chromosome X has 12 transcripts that are differentially expressed between the sexes (using a q value threshold of 0.05), four of which correspond to isoforms of known genes (XIST, KDM6A, TSIX, PNPLA4 and RPS4X). At the gene level, chromosome X has seven differentially expressed genes at the same q value cutoff.

17. Adds touch of tropical flair to our visualizations, making them visually appealing and thematic

```{r}
# Define a tropical color palette
tropical = c('darkorange', 'dodgerblue', 'hotpink', 'limegreen', 'yellow')

# Set the color palette for plots
palette(tropical)

```

18. Show the distribution of gene abundances (measured as FPKM values) across samples, colored by sex.

```{r}
# Extract FPKM values from ballgown object
fpkm = texpr(bg_chrX, meas = "FPKM")

# Log2 transformation of FPKM values
fpkm = log2(fpkm + 1)

geuvadis_phenodata$sex = as.factor(geuvadis_phenodata$sex)

# Create a boxplot with colors representing different sexes
Figure_1 <- boxplot(fpkm, col = geuvadis_phenodata$sex, las = 2, ylab = 'log2(FPKM+1)')

# Add title
title(main = "Distribution of FPKM values across the 12 samples")

# Initialize the PNG device
png("Distribution of FPKM values across the 12 samples.png")

# Save the base R boxplot as a figure
dev.copy(png, "Distribution of FPKM values across the 12 samples.png")

dev.off()

```

```{r}
# Create a data frame with 'sex' and 'fpkm' columns
plot_data <- data.frame(sex = geuvadis_phenodata$sex, fpkm = fpkm[12,])

# Create a new column named "Sample_id" with the values from the row names from plot_data
plot_data$Sample_id <- rownames(plot_data)

# Remove "FPKM." from the "Sample_id" column using sub function
plot_data$Sample_id <- sub("FPKM\\.", "", plot_data$Sample_id)

# Remove the original row names
rownames(plot_data) <- NULL

# Create a boxplot with colors representing different sexes
Figure_2 <- ggplot(plot_data, aes(x = sex, y = fpkm, fill = sex)) +
  geom_boxplot() +
  labs(title = "Distribution of Gene Abundances",
       x = "Sex",
       y = "FPKM Values") +
  scale_fill_manual(values = c('darkorange', 'dodgerblue')) +
  
  # Scale x-axis 
  scale_x_discrete(name = "Gender",
                   labels = c("Female", "Male")) +
  
  # Scale y-axis
  scale_y_continuous(name = "Fragments Per Kilobase Million",
                     breaks = seq(0, 5, by = 0.2))+
  
  # Remove legend to create a cleaner appearance
  theme(legend.position = "none")+

  # Customize axis labels
  theme(
    axis.title.x = element_text(size = 13, color = "Red", family = "mono"),
    axis.title.y = element_text(size = 13, color = "black", family = "mono"),
    panel.grid.major.x = element_line(colour = "darkgreen"),
    panel.grid.minor.x = element_line(colour = "darkgreen"),
    panel.grid.major.y = element_line(colour = "darkgreen"),
    panel.grid.minor.y = element_line(colour = "darkgreen"),
    plot.background = element_rect(fill = "tan2"),
    panel.background = element_rect(fill = "tan")
  )


# Save the boxplot as an image file
ggsave("Gene_abundance_boxplot.png", Figure_2)

# Display the boxplot
print(Figure_2)

```


19. Make plots of individual transcripts across samples. 

```{r}
# Get the name of the 12th transcript in the ballgown object (bg_chrX)
transcript_name <- ballgown::transcriptNames(bg_chrX)[11]

# Get the name of the gene associated with the 12th transcript
gene_name <- ballgown::geneNames(bg_chrX)[11]

# Create a ggplot scatterplot with jittered points
Figure_3 <- ggplot(plot_data, aes(x = sex, y = fpkm, color = sex)) +
  geom_boxplot(width = 0.5) +
  geom_jitter(width = 0.1, height = 0, size = 3) +
  labs(title = paste(gene_name, ":", transcript_name),
       x = "Sex",
       y = "log2(FPKM+1)",
       color = "Sex") +
  # Scale x-axis 
  scale_x_discrete(name = "Gender",
                   labels = c("Female", "Male")) +
  
  # Scale y-axis
  scale_y_continuous(name = "Fragments Per Kilobase Million",
                     breaks = seq(0, 5, by = 0.2))+
  
  # Remove legend to create a cleaner appearance
  theme(legend.position = "none")+

  
  # Customize axis labels
  theme(
    axis.title.x = element_text(size = 13, color = "Red", family = "mono"),
    axis.title.y = element_text(size = 13, color = "black", family = "mono"),
    panel.grid.major.x = element_line(colour = "darkgreen"),
    panel.grid.minor.x = element_line(colour = "darkgreen"),
    panel.grid.major.y = element_line(colour = "darkgreen"),
    panel.grid.minor.y = element_line(colour = "darkgreen"),
    plot.background = element_rect(fill = "tan2"),
    panel.background = element_rect(fill = "tan")
  )

# Display the plot
# Increase the width of the second column (Figure_3)
FPKM_distributions_in_males_and_females <- plot_grid(Figure_2, Figure_3, labels = c("A", "B"))


FPKM_distributions_in_males_and_females

```

```{r}
# Extract the FPKM values for the specified transcript and gene

transcript_index <- which(ballgown::transcriptNames(bg_chrX) == "NM_012227")
gene_index <- which(ballgown::geneNames(bg_chrX) == "GTPBP6")
fpkm_values <- texpr(bg_chrX)[transcript_index, ]

# Create a data frame with relevant data
plot_data <- data.frame(
  sex = geuvadis_phenodata$sex,
  fpkm = fpkm_values,
  gene = ballgown::geneNames(bg_chrX)[gene_index],
  transcript = ballgown::transcriptNames(bg_chrX)[transcript_index]
)

# Filter data for the specified gene and transcript
filtered_data <- subset(plot_data, gene == "GTPBP6" & transcript == "NM_012227")

# Create the box-and-whisker plot
Specified_transcript_and_gene <- ggplot(filtered_data, aes(x = sex, y = fpkm, fill = sex)) +
  geom_boxplot() +
  labs(title = paste("FPKM Distributions for Transcript NM_012227 (Gene GTPBP6)"),
       x = "Sex", y = "log2(FPKM+1)",
       fill = "Sex") +
  theme_minimal()

Specified_transcript_and_gene
```

20.Visualization of the assembled transcripts is done with the plotTranscripts function. Transcripts or exons can be colored by expression level. This plot colors transcripts by expression level

```{r}
# Plot transcripts for a specific gene in sample any of the sample

gene_index <- 1729
gene_id <- ballgown::geneIDs(bg_chrX)[gene_index]

# Create a PNG file with specified dimensions
png(file = "Visualization of the assembled transcripts.png", width = 1200, height = 600)

# Plot transcripts for two samples
plotTranscripts(gene_id, bg_chrX, 
  samples = c('ERR188044', 'ERR188104', 'ERR188234', 'ERR188245', 'ERR188257', 'ERR188273', 'ERR188337','ERR188383', 'ERR188401', 'ERR188428', 'ERR188454', 'ERR204916'),
  colorby = "transcript", meas = "FPKM", legend = TRUE, labelTranscripts = FALSE,
  main = paste("Gene ", gene_id, " Expression across the diffrent Samples"),
  blackBorders = TRUE, log = FALSE, logbase = 2, customCol = NULL, customOrder = NULL)

# Save the plot
dev.off()

```
Sometimes several very similar transcripts are assembled for the same gene, which might cause expression estimates for those transcripts to be unreliable: statistically, it can very difficult or impossible to tell which of two very similar transcript a read came from. This means differential expression results might also be unreliable. As a preliminary attempt at addressing this issue, Ballgown provides some simple transcript clustering functions. The idea is that similar assembled transcripts can be grouped together in clusters, and differential expression analysis could be performed on the cluster, whose expression measurement aggregates the expression estimates of the transcripts that compose it.

```{r}
# group a gene’s assembled transcripts into clusters
clusterTranscripts(gene_id, bg_chrX, k = 2, method = c("kmeans"))

#visualize the transcript clusters:
plotLatentTranscripts(gene_id, bg_chrX, k=2, method='kmeans', returncluster=FALSE)

```


21. We can also plot the average expression levels for all transcripts of a gene within different groups using the plotMeans function. We need to specify which gene to plot, which Ballgown object to use and which variable to group by. As an example, plot the gene MSTRG.56, using the following command.

```{r}
# Define the gene whose transcripts will be plotted.
Gene_id <- 'MSTRG.56'

# Create a plot of means grouped by 'sex'
plotMeans(Gene_id, bg_chrX_filt, groupvar = "sex", colorby = c("transcript"), legend = FALSE)

```


```{r}
# Create an Volcano plot
EnhancedVolcano(
  results_genes,
  lab = rownames(results_genes),
  x = 'fc',
  y = 'pval',
  xlim = c(-4, 4),
  ylim = (c(0, 10)),
  title = "Volcano Plot (p-value < 0.05)")
  
```



```{r}
# Assuming bg_chrX_filt is your Ballgown object
expression_matrix <- as.matrix(texpr(bg_chrX_filt))

# Get the row names of the top N genes based on maximum expression values
top_genes_ids <- names(sort(apply(expression_matrix, 1, max), decreasing = TRUE)[1:10])

# Extract expression data for the top N genes
top_expression <- expression_matrix[top_genes_ids, ]

# Create a heatmap using the numeric expression matrix of top genes
heatmap(top_expression, 
        Colv = NA,    # To disable hierarchical clustering of columns
        Rowv = NA,    # To disable hierarchical clustering of rows
        scale = "row", # Scale rows (genes)
        main = paste("Highly Expressed Genes Heatmap")
)
```

```{r}
# Extract gene names corresponding to the top gene IDs
matching_gene_names <- results_transcripts$geneIDs[results_transcripts$id %in% top_genes_ids]

# Extract expression data for the specified genes
expression_data_to_plot <- expression_matrix[top_genes_ids, ]

# Create a heatmap using the numeric expression matrix of specified genes
heatmap(expression_data_to_plot, 
        Colv = NA,    # To disable hierarchical clustering of columns
        Rowv = NA,    # To disable hierarchical clustering of rows
        scale = "row", # Scale rows (genes)
        main = "Selected Genes Heatmap",
        labRow = matching_gene_names  # Use matching gene names as row labels
)

```



```{r}

# Get the row names of the top N genes based on maximum expression values
top_genes_ids <- names(sort(apply(expression_matrix, 1, max), decreasing = TRUE)[1:20])

# Extract gene names corresponding to the top gene IDs
matching_gene_names <- results_transcripts$geneIDs[results_transcripts$id %in% top_genes_ids]

# Extract expression data for the specified genes
expression_data_to_plot <- expression_matrix[top_genes_ids, ]

# Set up colors for the gender annotations
colors <- c("Blue", "Pink") # Blue for Male and Pink for female

# Expand 'anno_color' to match the number of columns in 'heatmap_data'
anno_color_expanded <- data.frame(Annotation = matching_gene_names)

# Set up colors for the annotations
colors <- c("Blue", "Pink")

# Creating a heatmap with the expanded 'anno_color' and specified colors
pheatmap(expression_data_to_plot,
         annotation_col = anno_color_expanded,
         annotation_colors = list(Annotation = colors),
         scale = "row",
         main = "20 most variable attributes",
         cutree_rows = 2,
         cutree_cols = 3)

```

