---
title: "Introduction and Wrangling"
author: "Vikas Kauhsik"
date: "2023"
output:
  pdf_document: default
  html_document: default
---

## Set-Up

The dataset can be accessed here: https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+%28Diagnostic%29

The phenotype attributions are as follows: 

1) ID number
2) Diagnosis (M = malignant, B = benign)
3-32)

Ten real-valued features are computed for each cell nucleus:

a) radius (mean of distances from center to points on the perimeter)
b) texture (standard deviation of gray-scale values)
c) perimeter
d) area
e) smoothness (local variation in radius lengths)
f) compactness (perimeter^2 / area - 1.0)
g) concavity (severity of concave portions of the contour)
h) concave points (number of concave portions of the contour)
i) symmetry
j) fractal dimension ("coastline approximation" - 1)

Here is some code to read in the dataset

Use the `read.csv()` function to store the data from the `breastCancer.csv` file into an object called `bCancer` and check the data.

```{r}
# Read the CSV file "breastCancer.csv" into the 'bCancer' object with headers
bCancer <- read.csv("breastCancer.csv", header = TRUE)

# Get the dimensions (number of rows and columns) of the 'bCancer' object
dim(bCancer)

# Retrieve the column names of the 'bCancer' object
colnames(bCancer)

# Generate summary statistics for the 'bCancer' object
summary(bCancer)

# Display the first six rows of the 'bCancer' object
head(bCancer)

```
## Questions

1. When installing a package with the `install.packages()` function, the incoming package name needs to be enclosed in quotations. Why is this?

In R, text data (package names) is typically represented as character strings. The install.packages() function expects the package name as an argument. `The quotations is necessary to indicate that it's a string argument.` 

```{r}
# install.packages("tidyverse")
library(tidyverse)

```

2. Create a vector object with a numeric value, a character value, and a logical value. What data type is this vector and why? 
```{r}
#vector object with a numeric value, a character value, and a logical value
vector_object <- c(21092023, "September", TRUE)

#find the class of the vecotr object
class(vector_object)

```
This vector_object is of character data type due to the presence of the character value "September" in the vector. `Character takes priority, because it is the most “malleable” class. This is followed by numeric, then logical`. When combining different data types in a vector, R will select the data type that is highest in this hierarchy to avoid data loss.

3. Create a new object called `recent_fruits`. Create 5 rows with the rownames as different fruits and 3 columns named "color", "shape", and "taste". Describe the fruit's color and shape in one word, rank the taste out of 10. Fill out the table! 
```{r}
# Create a data frame with 3 columns named "Color", "Shape", and "Taste".
Color = c("Red", "Yellow", "Green", "Orange", "Purple")
Shape = c("Round", "Oval", "Cylindrical", "Spherical", "Irregular")
Taste = c(6, 9, 5, 8, 10)

recent_fruits <- data.frame(Color, Shape, Taste)

# Assign rownames to the data frame
Fruits <- c("Apple", "Banana", "Cucumber", "Orange", "Grapes")
rownames(recent_fruits) <- Fruits

# Print the data frame
print(recent_fruits)

```

4. Calculate the mean taste rank from your previous table. 
```{r}
# Calculate the mean taste rank
mean_taste_rank <- mean(recent_fruits$Taste)

# Print the mean taste rank
print(mean_taste_rank)

```


5. From the bCancer dataset, subset the samples with a mean radius less than 20 into a new object called `large_tumor`

```{r}
colnames(bCancer)

#Subset samples with a mean radius less than 20
large_tumor <- filter(bCancer, radius_mean < 20)

##display the first six rows of the new object
head(large_tumor)
         
```
6. From the `large_tumor` object, subset for the columns describing the `id`, `diagnosis`, `radius_mean`, `texture_mean`, `smoothness_mean`, and `compactness_mean`. Reshape the data into a longer format so that there are only three columns: id, variable, and value.


```{r}
large_tumor_subset <- large_tumor %>% 
#selects specific columns from the large_tumor object
select(id, diagnosis, radius_mean, texture_mean, smoothness_mean, compactness_mean) %>%
  
#Reshapes the data from a wide format to a long format. It takes the allocated columns and gathers them into variable and value column.
mutate(diagnosis = parse_number(diagnosis)) %>% 
pivot_longer(cols = c(diagnosis, radius_mean, texture_mean, smoothness_mean,compactness_mean), 
             names_to =  "variable", values_to = "value") %>% 
  
# Remove rows with NA values from the object 
na.omit(large_tumor)
  
#display the first six rows of the reshaped object 
head(large_tumor_subset)
```
The warning message indicates that there are 524 parsing failures while trying to convert values in the diagnosis column to numeric using parse_number(). The issue is due to the fact that the diagnosis column contains non-numeric character values. 






