---
title: "Visualization"
author: "Vikas kaushik"
date: "2023"
output:
  pdf_document: default
  html_document: default
---

## Data Visualization
```{r}
# get the path of the current working directory
getwd()

# Call all the important libraries for this assignment
library (tidyverse)
library (cowplot)
library (datasauRus)

```

The dataset can be accessed here: https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+%28Diagnostic%29

The phenotype attributions are as follows: 

1) ID number
2) Diagnosis (M = malignant, B = benign)
3-32)

Ten real-valued features are computed for each cell nucleus:

a) radius (mean of distances from center to points on the perimeter)
b) texture (standard deviation of gray-scale values)
c) perimeter
d) area
e) smoothness (local variation in radius lengths)
f) compactness (perimeter^2 / area - 1.0)
g) concavity (severity of concave portions of the contour)
h) concave points (number of concave portions of the contour)
i) symmetry
j) fractal dimension ("coastline approximation" - 1)

Here is some code to read in the dataset

```{r}
# Read the CSV file "breastCancer.csv" into a object bCancer
bCancer <- read.csv("breastCancer.csv", header = T, stringsAsFactors = T)

# Display the structure of the bCancer data frame
str(bCancer)

# Provide a summary of the bCancer data frame
summary(bCancer)

# Display the first 5 rows of the bCancer data frame
head(bCancer, 5)

```

1. The aesthetic of the plot can be defined in both the main `ggplot()` function or the plot specific geometry function. What advantage does this redundancy allow for? Give an example of this use case. 

Ans`Redundancy in defining aesthetics at both the main ggplot() level and the specific geometry (geom_*()) level provides a balance between global consistency and local customization. Global aesthetics set in ggplot() apply universally to all layers, ensuring a cohesive look, while local aesthetics in geom_*() allow for specific adjustments or highlights within individual layers. This flexibility is particularly valuable in creating complex and informative visualizations.`Example of this use case:

```{r}

# Defining aesthetics in ggplot() for global settings
plot1 <- ggplot(bCancer, aes(x = radius_mean, y = concavity_mean, color = diagnosis)) +

  # First layer: Scatter plot with specific aesthetics
  geom_point(size = 1, alpha = 0.8) +
  
  # Second layer: Line plot with different aesthetics
  geom_line(size = 1, linetype = "dashed") +
  
  # Scale y axis 
  scale_y_continuous(name = "Concavity Mean",
                     breaks = seq(0, 0.5, by = 0.1))+
  
  # Global title and axis labels
  labs(title = "Scatter and Line Plot", x = "Radius Mean")

  
# Global aesthetics set in ggplot()
plot2 <- ggplot(bCancer, aes(x = radius_mean, y = concavity_mean)) +
  
  # Layer 1: Scatter plot with local aesthetics
  geom_point(color = "red", size = 2, alpha = 0.1) +
  
  # Layer 2: Line plot with different local aesthetics
  geom_line(color = "black", linetype = "dashed", size = 0.5) +
  
  # Global labels and theme
  labs(title = "Scatter and Line Plot", y = "Concavity Mean")+
  
  # Scale x axis
  scale_x_continuous(name = "Radius Mean",
                     breaks = seq(0, 30, by = 3))

Defining_aesthetics <- plot_grid(plot1, plot2, labels = c("A", "B"), ncol = 1)

Defining_aesthetics

```

2. What is the difference between a bar plot and a histogram? Which is more appropriate when I am plotting the heights of all students in this course?

Ans`Histograms are employed when dealing with continuous data, usually numeric, to visualize the distribution of values across different bins or intervals. Bar charts, on the other hand, are suited for categorical data, where each bar represents a distinct category or group. In the context of plotting the heights of all students, I would use a histogram because height is a continuous variable, and a histogram allows me to observe the distribution of heights. A bar chart, designed for categorical data, wouldn't be the most suitable choice for visualizing the distribution of continuous variables like heights.`


```{r}
# Set up the basic ggplot object with data and aesthetic mapping
Bar <- ggplot(bCancer, aes(x = diagnosis, fill = diagnosis)) +
  
  # Add the bar chart layer with custom aesthetics
  geom_bar(alpha = 0.7) +
  
  # #Customize colors "red" is for malignant, and "green" is for benign.
  scale_fill_manual(values = c("green", "red")) +
  
  # Add labels and title to the plot
  labs(title = "Bar Plot", x = "Diagnosis", y = "Count") +
  
  # Scale x-axis
  scale_x_discrete(name = "Diagnosis",
                   labels = c("Benign", "Malignant"))


# Set up the basic ggplot object with data and aesthetic mapping
Hist <- ggplot(bCancer, aes(x = radius_mean, fill = diagnosis)) +
  
  # Add the histogram layer with custom aesthetics
  geom_histogram(binwidth = 5, alpha = 0.7) +
  
  #Customize colors "red" is for malignant, and "green" is for benign.
  scale_fill_manual(values = c("green", "red")) +
  
  # Add labels and title to the plot
  labs(title = "Histogram", x = "Radius_mean", y = "Frequency")+
  
  # Scale x axis
  scale_x_continuous(name = "Radius Mean",
                     breaks = seq(0, 30, by = 5))

# Set up the basic ggplot object with data and aesthetic mapping
Box <- ggplot(bCancer, aes(x = diagnosis, y = radius_mean, fill = diagnosis)) +

  # Add the boxplot layer
  geom_boxplot() +
  
  # Add labels and title to the plot
  labs(title = "Boxplot", y = "Radius Mean")+
  
  # #Customize colors "red" is for malignant, and "green" is for benign.
  scale_fill_manual(values = c("green", "red")) +
  
  # Scale x-axis
  scale_x_discrete(name = "Diagnosis",
                   labels = c("Benign", "Malignant"))+
  
  # Remove legend to create a cleaner appearance
  theme(legend.position = "none")+

  # Add horizontal lines at y = 6.5 and y = 8.5
  geom_hline(yintercept = 6.5) +
  geom_hline(yintercept = 8.5)

# Set up the basic ggplot object with data and aesthetic mapping
Violin <- ggplot(bCancer, aes(x = diagnosis, y = radius_mean, fill = diagnosis)) +
  
  # Add the boxplot layer
  geom_violin() +
  
  # Add labels and title to the plot
  labs(title = "Violin Plot", y = "Radius Mean")+
  
  # #Customize colors "red" is for malignant, and "green" is for benign.
  scale_fill_manual(values = c("green", "red")) +
  
# Scale x-axis
  scale_x_discrete(name = "Diagnosis",
                   labels = c("Benign", "Malignant"))+
  
  # Remove legend to create a cleaner appearance
  theme(legend.position = "none")


# Arrange plots in a 2,2 grid
Bar_Hist_Box_Violin <- plot_grid(Bar, Hist, Box, Violin, labels = c("A", "B", "C", "D"), ncol = 2)

Bar_Hist_Box_Violin

```
3. When comparing the mean radius between diagnosis groups, what plot type would be most appropriate? Justify your response by relating back to the type of data we are portraying. Then, create the plot. Make sure to format the plot properly such as labeling the titles and axis. 

Ans`For comparing the mean radius between diagnosis groups (Malignant and Benign), suitable plot type is going to be grouped bar chart. Grouped bar charts are effective for visualizing and comparing means or other summary statistics across different categories`

```{r}
# Set up the basic ggplot object with data and aesthetic mapping
GBar <- ggplot(bCancer, aes(x = diagnosis, y = radius_mean, fill = diagnosis)) +
  
  # Add the bar chart layer with custom aesthetics
  geom_bar(alpha = 0.7, stat = "summary", fun = "mean") +
  
  # Customize colors "red" is for malignant, and "green" is for benign
  scale_fill_manual(values = c("green", "red")) +
  
  # Add labels and title to the plot
  labs(title = "Comparing the mean Radius") +
  
  # Scale x-axis
  scale_x_discrete(name = "Diagnosis",
                   labels = c("Benign", "Malignant"))+
  
  # Scale y axis ("Mean Radius" is continuous)
  scale_y_continuous(name = "Mean Radius",
                     breaks = seq(0, 20, by = 4))+
  
  # Remove legend to create a cleaner appearance
  theme(legend.position = "none")
  

# Set up the basic ggplot object with data and aesthetic mapping
Jitter <- ggplot(bCancer, aes(x = diagnosis, y = radius_mean, colour = diagnosis)) +
  
  # Add jitter points for each data point with custom parameters
  geom_jitter(
    width = 0.3,      # Adjust the width of the jitter
    height = 0.05,    # Adjust the height of the jitter
    shape = 16,       # Set the shape of the points (16 for circles)
    alpha = 0.7       # Set the transparency of the points
  ) +
 
  # Customize colors "red" is for malignant, and "green" is for benign
  scale_colour_manual(values = c("green", "red")) + 
  
  # Add labels and title to the plot
  labs(title = "Comparing the mean Radius") +
  
  # Scale x-axis
  scale_x_discrete(name = "Diagnosis",
                   labels = c("Benign", "Malignant")) +
  
  # Scale y-axis ("Mean Radius" is continuous)
  scale_y_continuous(name = "Mean Radius",
                     breaks = seq(0, 30, by = 4))+
  
  # Remove legend to create a cleaner appearance
  theme(legend.position = "none")+
  
  # Theme customization
  theme(
    axis.title.x = element_text(family = "mono", color = "red", size = 12),
    axis.title.y = element_text(family = "serif", color = "red", size = 10),
    axis.line.x = element_line(linewidth = 1, color = "blue"),
    axis.line.y = element_line(linewidth = 1, color = "blue"),
    axis.ticks.y = element_line(linewidth = 5),
    axis.ticks.length.y = unit(-.2, "cm"),
    panel.grid.major = element_line(colour = "lightgreen"),
    panel.grid.minor.y = element_line(colour = "darkgreen"),
    plot.background = element_rect(fill = "tan"),
    panel.background = element_rect(fill = "tan2")
  )
  
# Arrange plots in a 1,2 grid
GBar_Jitter <- plot_grid(GBar, Jitter, labels = c("A", "B"))
GBar_Jitter

```

4. When visualizing if there is a relationship between the mean smoothness and mean compactness of the tumors, what plot type would be most appropriate? Justify your response by relating back to the type of data we are portraying. Then, create the plot. Make sure to format the plot properly such as labeling the titles and axis. 

Ans`To visualize the relationship between the mean smoothness and mean compactness of tumors, a scatter plot would be most appropriate. Scatter plots are effective for displaying the correlation or relationship between two continuous variables.`

```{r}
# Scatter plot for mean smoothness vs mean compactness
ScatterPlot <- ggplot(bCancer, aes(x = smoothness_mean, y = compactness_mean, color = diagnosis)) +
  
  # Add points for each data point
  geom_point(alpha = 0.7) +
  
  # Customize colors "red" is for malignant, and "green" is for benign
  scale_color_manual(values = c("green", "red")) +
  
  # Add labels and title to the plot
  labs(title = "Relationship Between Mean Smoothness and Mean Compactness") +
   
  # Scale x-axis 
  scale_x_continuous(name = "Mean Smoothness",
                     breaks = seq(0, 0.2, by = .025))+
  
  # Scale y-axis
  scale_y_continuous(name = "Mean Compactness",
                     breaks = seq(0, 0.5, by = 0.1))+
  
  # Remove legend to create a cleaner appearance
  theme(legend.position = "none")+
  
  # Draw a correlation line using linear regression to show relationship
  geom_smooth(method = "lm", se = FALSE, linetype = "solid", color = "blue") +
  
  # Customize axis labels
  theme(
    axis.title.x = element_text(size = 13, color = "black", family = "mono"),
    axis.title.y = element_text(size = 13, color = "black", family = "mono"),
    axis.line.x = element_line(linewidth = 1, color = "black"),
    axis.line.y = element_line(linewidth = 1, color = "black"),
    axis.ticks.y = element_line(linewidth = 7),
    axis.ticks.length.y = unit(-.2, "cm"),
    axis.ticks.x = element_line(linewidth = 7),
    axis.ticks.length.x = unit(-.2, "cm"),
    panel.grid.major.x = element_line(colour = "darkgreen"),
    panel.grid.minor.x = element_line(colour = "darkgreen"),
    panel.grid.major.y = element_line(colour = "darkgreen"),
    panel.grid.minor.y = element_line(colour = "darkgreen"),
    plot.background = element_rect(fill = "tan2"),
    panel.background = element_rect(fill = "tan")
  )

# Display the scatter plot
print(ScatterPlot)

```

```{r}
# Save figure 1
ggsave("Defining_aesthetics.png", Defining_aesthetics, height = 12, width = 14)

# Save figure 2
ggsave("Bar_Hist_Box_Violin.png", Bar_Hist_Box_Violin, height = 12, width = 12)

# Save figure 3
ggsave("GBar_Jitter.png", GBar_Jitter, height = 10, width = 16)

# Save figure 3
ggsave("ScatterPlot.png", ScatterPlot, height = 6, width = 8)

```

